---
title: "Quick Guide to XML in R"
author: "Jesse Lecy"
date: "July 7, 2016"
output: html_document
---



The IRS has released data on 990 tax returns from electronic nonprofit and foundation filers. The data is reported as XML documents for each return. 

In order to make the data accessible, we are working to 



## BASIC DATA PRINCIPLES

Data used for statistical analysis often comes in flat, two-dimensional spreadsheets. Rows represent observations, and columns represent variables. 

Out in the real world, however, data is often generated through relational databases that can capture the one-to-many relationships inherent in systems.

As an example, if you have shopped on Amazon your customer information has been stored in their database, and information about each transaction has been captured. They use two different tables for this information - one for customers and one for transactions. If they tried to store all of this information in a single spreadsheet much of the static information (your name and address) would have to be repeated. Since they have billions of transactions, each with hundreds of fields, that spreadsheet would quickly become large and slow (computationally expensive, as they say in the industry).

Relational databases - storing information that has a one-to-many relationship in different tables - simplifies life and makes for robust systems. If the information needs to be combined, for example for market analysis to look at the relationship between products and zip codes, it can be done through a query. 


```{r, echo=FALSE}

customer.info <- data.frame(
                    CUSTOMER.ID=c("178","934","269"),
                    FIRST.NAME=c("Alvaro","Janette","Latisha"),
                    LAST.NAME=c("Jaurez","Johnson","Shane"),
                    ADDRESS=c("123 Park Ave","456 Candy Ln","1600 Penn Ave"),
                    ZIP.CODE=c("57701","57701","20500"))

purchases <- data.frame(
                    CUSTOMER.ID=c("178","178","269","269","934"),
                    PRODUCT=c("video","shovel","book","purse","mirror"),
                    PRICE=c(5.38,12.00,3.99,8.00,7.64) )

```

```{r}
customer.info

purchases

merge( customer.info, purchases )

```




## XML DATA MODELS

The world of web programming requires a different way of representing data. If we want to build smart websites, browsers and web applications need to be able to make sense of data. As a result, eXtensible Markup Langauge was invented in order to (1) provide context for informaton so it can be made more useful, and (2) provide a flat, linear representation of relational databases. 

XML works by adding tags to information so that computers can make sense of it:

```
<ANIMAL>DOG</ANIMAL>

<PLANT>FLOWER</PLANT>

```

This scheme also allows for the creation of sets of things.


```
<ANIMALS>

   <FISH>Tuna</FISH>
   <MAMMAL>Dog</MAMMAL>
   <SNAKE>Viper</SNAKE>
   
</ANIMALS>

```

We can now describe the world in a couple of different ways - we can ask for the set of all animals, or we can ask for specific types of animals. 

Similarly, here is what part of the relational database above would look like.


```
<MEMBERS>

   <CUSTOMER>
      <ID>178</ID>
      <FIRST.NAME>Alvaro</FIRST.NAME>
      <LAST.NAME>Juarez</LAST.NAME>
      <ADDRESS>123 Park Ave</ADDRESS>
      <ZIP>57701</ZIP>
   </CUSTOMER>
   
   <CUSTOMER>
      <ID>934</ID>
      <FIRST.NAME>Janette</FIRST.NAME>
      <LAST.NAME>Johnson</LAST.NAME>
      <ADDRESS>456 Candy Ln</ADDRESS>
      <ZIP>57701</ZIP>
   </CUSTOMER>  
   
</MEMBERS>

<PURCHASES>

   <TRANSCTION>
      <ID>178</ID>
      <PRODUCT>video</PRODUCT>
      <PRICE>5.38</PRICE>
   </TRANSACTION>
   
   <TRANSCTION>
      <ID>178</ID>
      <PRODUCT>shovel</PRODUCT>
      <PRICE>12.00</PRICE>
   </TRANSACTION>
   
</PURCHAES>

```






## ACCESSING DATA WITH XML

Accessing data in the XML format requires a basic understanding of two principles - nodes and paths. 

Let's return to the example above:

```
<MEMBERS>

   <CUSTOMER>
      <ID>178</ID>
      <FIRST.NAME>Alvaro</FIRST.NAME>
      <LAST.NAME>Juarez</LAST.NAME>
      <ADDRESS>123 Park Ave</ADDRESS>
      <ZIP>57701</ZIP>
   </CUSTOMER>
   
   <CUSTOMER>
      <ID>934</ID>
      <FIRST.NAME>Janette</FIRST.NAME>
      <LAST.NAME>Johnson</LAST.NAME>
      <ADDRESS>456 Candy Ln</ADDRESS>
      <ZIP>57701</ZIP>
   </CUSTOMER>  
   
</MEMBERS>
```

Here each tag represents a separate node, so this XML document contains three layers of nodes - members, customers, and the set of nodes related to customer information.

You can start to see that the nested nature of data creates a tree structure. Since it is hierarchical, each **parent** node contains **children** nodes. MEMBERS contains CUSTOMER nodes. CUSTOMER contains ID, NAME, and ADDRESS nodes. 

To manipulate the data, we can grab a single node at once, and work with all of its sub-elements. 

We will use the *xml2* package and the *xml_parent()* and *xml_children()* functions. The function *xml_name()* prints the name of the node.

```{r}

# install.packages( "xlm2" )
# install.packages( "dplyr" )

library( xml2 )
library( dplyr )

dat <- read_xml( "<MEMBERS>
                     <CUSTOMER>
                        <ID>178</ID>
                        <FIRST.NAME>Alvaro</FIRST.NAME>
                        <LAST.NAME>Juarez</LAST.NAME>
                        <ADDRESS>123 Park Ave</ADDRESS>
                        <ZIP>57701</ZIP>
                     </CUSTOMER>
                     <CUSTOMER>
                        <ID>934</ID>
                        <FIRST.NAME>Janette</FIRST.NAME>
                        <LAST.NAME>Johnson</LAST.NAME>
                        <ADDRESS>456 Candy Ln</ADDRESS>
                        <ZIP>57701</ZIP>
                     </CUSTOMER>  
                  </MEMBERS>"  )


xml_name(  dat  )

xml_name( xml_parent( dat ) )  # no parents - it is a root node

xml_name( xml_children( dat ) )

# print the full path directory:

dat %>% xml_find_all( '//*') %>% xml_path()

```



We can break the document apart into individual nodes, and treat them separately using *xml_find_first()* and *xml_find_all()*.

```{r}

customer1 <- xml_find_first( dat, "//CUSTOMER" )

xml_name(  customer1  )

xml_name( xml_parent( customer1 ) )

xml_name( xml_children( customer1 ) )


```


And finally, we can access data within a "leaf" node using the *xml_text()*, *xml_double()*, and *xml_integer()* commands. 


```{r}

xml_text( xml_find_first( dat, "//ADDRESS" ) )

xml_text( xml_find_all( dat, "//ADDRESS" ) )

```

Putting it together, we can start to build datasets:

```{r}

first.name <- xml_text( xml_find_all( dat, "//FIRST.NAME" ) )
last.name <- xml_text( xml_find_all( dat, "//LAST.NAME" ) )
address <- xml_text( xml_find_all( dat, "//ADDRESS" ) )
zip <- xml_text( xml_find_all( dat, "//ZIP" ) )

data.frame( first.name, last.name, address, zip )

```

## NAME SPACES

The header information in an XML document - the root node - may contain information on "name spaces". 

```

# contains no name spaces

<Return returnVersion="2014v5.0">
  
  
# contains name spaces

<Return xmlns="http://www.irs.gov/efile" xmlns:xsi="http://www.w3.org/2001/XMLSchema-instance" xsi:schemaLocation="http://www.irs.gov/efile" returnVersion="2014v5.0">
  
```

This topic is beyond this short intro to XML, other than to note you do not need them to access the data and they will give you a headache, so remove them using the *xml_ns_strip()* function.


```
# LOAD DATA FROM A PC FILING FULL 990

dat <- read_xml( x="https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml", options=NULL )

# STRIP THE NAMESPACE

xml_ns_strip( dat )

```



## XPATH

In the XML structure there is a specific convention for referencing nodes called *xpath*. You can create sophisticated statements to access specific elements of your data, but the basic principles of a search statement are:

* Place two backslashes in front of the top node in your search
* Separate subsequent nodes with a single backslash
* Add a period if you want the search to be local

Node that arguments to the search functions *xml_find_all()* and *xml_find_first()* are a nodeset, and an xpath.

```{r}

xml_text( xml_find_first( dat, "//ADDRESS" ))

# if the ADDRESS field name is used in other parts of the data
# you will need to be more specific

xml_text( xml_find_first( dat, "//CUSTOMER/ADDRESS" ))

xml_text( xml_find_first( dat, "//MEMBERS/CUSTOMER/ADDRESS" ))


# drilling down into a node

xml_text( xml_find_all( dat, "//ADDRESS" )) # returns all customer addresses

customer1 <- xml_find_first( dat, "//CUSTOMER" )  # create node for first customer only

xml_text( xml_find_all( customer1, ".//ADDRESS" ))

# note the use of the period in the xpath!

xml_text( xml_find_all( customer1, ".//ADDRESS" )) # with period search is local to nodeset

xml_text( xml_find_all( customer1, "//ADDRESS" )) # no period still searches full document


```

You can identify specific paths using the *xml_path()* function. This will give you a way to reference each non-unique node in the path individually.

```{r}


xml_path( xml_find_all( dat, "//ADDRESS") )

xml_text( xml_find_all( dat, "//MEMBERS/CUSTOMER[1]/ADDRESS" ))

xml_text( xml_find_all( dat, "//MEMBERS/CUSTOMER[2]/ADDRESS" ))

```


## XML ATTRIBUTES

Some nodes contain attributes that store meta-data associated with the node:

```
# contains no attributes

<ReturnData> ... </ReturnData>

<IRS990> ... </IRS990>

# contains attributes

<ReturnData documentCnt="6"> ... </ReturnData>

<IRS990 referenceDocumentId="RetDoc1044400001"> ... </IRS990>

```

We can access meta-data using the *xml_attr()* function.


```{r}

# LOAD DATA FROM A PC FILING FULL 990

dat <- read_xml( x="https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml", options=NULL )

# STRIP THE NAMESPACE

xml_ns_strip( dat )


# GRAB THE ATTRIBUTES

return.data <- xml_find_first( dat, "//ReturnData" )  # grab return data node

xml_attrs( x=return.data )  # list attributes for node

xml_attr( x=return.data, attr="documentCnt" )  # return document count attribute


```



## 990 DATA

Now for applying these rules to the data at hand. If we look at the structure of a 990 return, it looks something like this:



```

990 RETURN

   HEADER DATA
   
      NAME, EIN, YEAR, etc.
      
   RETURN DATA
   
      REVENUES
      EXPENSES
      GOVERNANCE
      
        BOARD MEMBERS
      
      FUNCTIONAL REVENUE CATEGORIES
      FUNCTIONAL EXPENSE CATEGORIES
      
      SCHEDULES
      
        A-Public Support
        B-Contributors
        D-Supplemental Financial Statements
        M-Non-Cash Contributions
        O-Supplemental Information
        R-Related Organizations
        
END 990 RETURN


```

Note that the structure varies between the 990, 990-EZ, and 990-PF returns, and that schedules included will vary by organization and year. 

If you would like to see an example of the full structure, run this code (not executed here because the output is long):

```{r}

# dat <- read_xml( "https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml" )
# xml_ns_strip( dat )
# dat %>% xml_find_all( '//*') %>% xml_path()

```

Here is the example organization listed on on the IRS Amazon Web Server page:

https://aws.amazon.com/public-data-sets/irs-990/

```{r}


# LOAD DATA FROM A PC FILING FULL 990

dat <- read_xml( x="https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml", options=NULL )

# STRIP THE NAMESPACE

xml_ns_strip( dat )



# EXAMINE THE DATA

# xml_structure( dat )  # this is overwhelming, we need to be able to drill down
 
xml_name( xml_root( dat ) )  # only one root node that contains all data

xml_name( xml_children( dat ) )  # split into two sections, header and return data

xml_name( xml_children( xml_find_first( dat, "//ReturnHeader" ) ) )

xml_name( xml_children( xml_find_first( dat, "//ReturnData/IRS990" ) ) )[ 1:10 ]

xml_text( xml_children( xml_find_first( dat, "//ReturnData/IRS990" ) ) )[ 1:10 ]


```

You can see that any node that contains children (for example USAddress) will print the data separated by a carriage return (\\n). 

As an example of how you might parse the data to create a dataset:

```{r}

header.data <- xml_find_first( dat, "//ReturnHeader" )


TaxYr <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/TaxYr" ) )
ReturnType <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/ReturnTypeCd" ) )
EIN <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/EIN" ) )
BusinessName <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/BusinessName/BusinessNameLine1Txt" ) )
Address <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/USAddress/AddressLine1Txt" ) )
City <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/USAddress/CityNm" ) )
State <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/USAddress/StateAbbreviationCd" ) )
Zip <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/Filer/USAddress/ZIPCd" ) )
TaxPrep <- xml_text( xml_find_all( dat, "//Return/ReturnHeader/BusinessOfficerGrp/DiscussWithPaidPreparerInd" ) )


header.df <- data.frame( TaxYr, ReturnType, EIN, BusinessName, Address, City, State, Zip, TaxPrep )

header.df

```

To build a full dataset, you will need to iterate over multiple 990 returns and combine these data. This would be approximately:

```

library( dplyr )

dat1 <- read_xml( url.01 )

# create header.df.01

dat2 <- read_xml( url.02 )

# create header.df.02

bind_rows( header.df.01, header.df.02 )

```


## Reconciling Fields Across Years

Some brilliant person or persons at the IRS have taken it upon themselves to change field names within their database system, so it is now not possible to easily scrape data across multiple years. Most change are minor:

**BusinessNameLine1** changed to **BusinessNameLine1Txt**

**FormType** changed to **FormTypeCd**

Unfortunately even small changes will break your code since it requires an exact match. As a result, you need to write the xpath statement in such a way that it will match either version. The queries have a bunch of 

```{r, eval=F}

library( xml2 )

# EXAMPLE 990 File
doc <- read_xml( "https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml" )
xml_ns_strip( dat )



# 2014 / 2015 990 VERSION
#
# "//Return/ReturnData/IRS990/GrossReceiptsAmt"


# 2013 / 2012 VERSION
#
# "//Return/ReturnData/IRS990/GrossReceipts"


# This works with 2014, 2015 data, fails with 2012, 2013 data

GROSSRECEIPTS <- xml_text( xml_find_all( doc, "//Return/ReturnData/IRS990/GrossReceiptsAmt" ) )


# INCLUSIVE XPATH VERSION WHEN TWO VARIABLE NAMES HAVE COMMON PREDICATE PATH

GROSSRECEIPTS <- xml_text( xml_find_all( doc, "/Return/ReturnData/IRS990/*[self::GrossReceipts or self::GrossReceiptsAmt]" ) )



# 990EZ CHANGES PATH STRUCTURE
#
# //Return/ReturnData/IRS990EZ/GrossReceiptsAmt



# NEW INCLUSIVE STATEMENT MUST uSE OR OPERATOR "|" AND FULL PATHS

gross.receipts.xpath <- "//Return/ReturnData/IRS990/GrossReceipts|//Return/ReturnData/IRS990/GrossReceiptsAmt|//Return/ReturnData/IRS990EZ/GrossReceiptsAmt"

GROSSRECEIPTS <- xml_text( xml_find_all( doc, gross.receipts.xpath ) )



```

Building a module to collect data from a portion of the 990s requires the user to specify all of the possible iterations of the xpath + field names for a given variable. If the variable is present on both the 990 and 990EZ that consists of two 990 version (pre and post 2014), and two 990EZ versions (pre and post 2014). This does not include 990-PF private foundations.

Here is an example of variation in field names between 2012 and 2014.

```
# 2012 990 VERSION


 [47] "/Return/ReturnData/IRS990/GrossReceipts"
 [48] "/Return/ReturnData/IRS990/GroupReturnForAffiliates"
 [49] "/Return/ReturnData/IRS990/Organization501c3"
 [50] "/Return/ReturnData/IRS990/WebSite"
 [51] "/Return/ReturnData/IRS990/TypeOfOrganizationCorporation"
 [52] "/Return/ReturnData/IRS990/YearFormation"
 [53] "/Return/ReturnData/IRS990/StateLegalDomicile"
 [54] "/Return/ReturnData/IRS990/ActivityOrMissionDescription"
 [55] "/Return/ReturnData/IRS990/NbrVotingMembersGoverningBody"
 [56] "/Return/ReturnData/IRS990/NbrIndependentVotingMembers"
 [57] "/Return/ReturnData/IRS990/TotalNbrEmployees"
 [58] "/Return/ReturnData/IRS990/TotalNbrVolunteers"
 [59] "/Return/ReturnData/IRS990/TotalGrossUBI"
 [60] "/Return/ReturnData/IRS990/NetUnrelatedBusinessTxblIncome"
 
 
 
 
 
# 2014 990 VERSION
 
 [47] "/Return/ReturnData/IRS990/GrossReceiptsAmt"
 [48] "/Return/ReturnData/IRS990/GroupReturnForAffiliatesInd"   
 [49] "/Return/ReturnData/IRS990/Organization501c3Ind"         
 [50] "/Return/ReturnData/IRS990/WebsiteAddressTxt"          
 [51] "/Return/ReturnData/IRS990/TypeOfOrganizationCorpInd"         
 [52] "/Return/ReturnData/IRS990/FormationYr"
 [53] "/Return/ReturnData/IRS990/LegalDomicileStateCd"
 [54] "/Return/ReturnData/IRS990/ActivityOrMissionDesc"
 [55] "/Return/ReturnData/IRS990/VotingMembersGoverningBodyCnt"
 [56] "/Return/ReturnData/IRS990/VotingMembersIndependentCnt"
 [57] "/Return/ReturnData/IRS990/TotalEmployeeCnt"
 [58] "/Return/ReturnData/IRS990/TotalVolunteersCnt"
 [59] "/Return/ReturnData/IRS990/TotalGrossUBIAmt"
 [60] "/Return/ReturnData/IRS990/NetUnrelatedBusTxblIncmAmt"
 
 
```



## Full Path Directory for Example 990 Return

To create a list of all variables and their respective xpaths available on a specific 990 return, you can use the following code: 

```{r}

dat <- read_xml( "https://s3.amazonaws.com/irs-form-990/201541349349307794_public.xml" )
xml_ns_strip( dat )

# NAMES OF FIELDS

dat %>% xml_find_all( '//*') %>% xml_name()

# FULL PATH FOR EASY REFERENCE WHEN NODES ARE NON-UNIQUE

dat %>% xml_find_all( '//*') %>% xml_path()

```

